---
title: "Intro to Time Series Wrangling & Viz"
author: "Sydney Rilum"
date: "1/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library(feasts) #ACF
library(slider)
```

## Read in the data

```{r}
toolik <- read_csv(here("data", "toolikweather.csv"))
```

### Convert this into a tsibble
(time series data frame)
need to convert data from a character to numeric class

```{r}
toolik_ts <- toolik %>% 
  mutate(date = lubridate::mdy(date)) %>%   #converts date to readable date form in R
  as_tsibble(key = NULL, index = date)  #sets what index is
```

```{r}
# shows observation for each day of year
ggplot(data = toolik_ts, aes(x = date, y = mean_airtemp)) +
  geom_line()
```

## Use `index_by()` function to aggregate time series information by intervals we specify

Aggregate by month:
```{r}
toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% #aggregate time series index by year and month
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE)) #calc mean of daily mean temp

#just like group_by, summarize

# shows observation for each month of year, smoother than previous graph
ggplot(data = toolik_month, aes(x = yr_mo, y = monthly_mean_temp)) +
  geom_line() +
  facet_wrap(~month(yr_mo, label = TRUE)) #pulls month out of yearmonth column
```

Aggregate by week:
```{r}
toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_mean_temp = mean(mean_airtemp, na.rm = TRUE),
            min_airtemp = min(mean_airtemp))

head(toolik_weekly) # see first few rows of observations
```

## Filtering time series

Use `filter_index()` to specify ranges of time series to keep or exclude.

```{r}
toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10")
```

```{r}
toolik_ts %>% 
  filter_index("2006-04-10" ~ "2006-05-15")
```

Look into different `filter_index()` functions!!!!

## Seasonplots

```{r}
toolik_ts %>% 
  filter(year(date) > 2014) %>%  #or could use `filter_index()`
  gg_season(y = mean_airtemp)

#season plots can show seasonality of data

toolik_month %>% 
  gg_season(y = monthly_mean_temp) +
  theme_light()
```

## Seasonal subseries plot
```{r}
toolik_month %>% 
  gg_subseries(monthly_mean_temp)
```

## Find moving average
Using the slider package in R

```{r}
set.seed(2021)
test <- rnorm(n = 100,
              mean = 40,
              sd = 10)

slide(test, ~.x, .before = 2, .after = 2) # makes a window of size 5 (includes 2 windows before and after)

# will convert this list to a numeric vector
```

```{r}
#rolling window, calculates moving average for windows of size 15 (days)
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp,
                                   mean,
                                   .before = 7,
                                   .after = 7))) #give it vector name, 7 obs before and after, calculate mean of all 15 obs, and returns value in new vector in new coloumn

ggplot() +
  geom_line(data = toolik_ts, aes(date, y = mean_airtemp), # raw data
            size = 0.2,
            color = "gray") +
  geom_line(data = roll_toolik_15, aes(x = date, y = ma_15d)) + # 15 days averages
  theme_minimal()

```

## Create the ACF (autocorrelation function)

```{r}
toolik_ts %>% 
  ACF(mean_airtemp) %>% 
  autoplot()
# lag in days, observations 1 day after are strongly positively correlated to each other

toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot()
# lag in months, observations that are 12 months apart (even 24 months!) tend to be strongly positively correlated, but observations that are 6 or 18 months apart (opposite seasons) see negative correlation
# air temps in 2010 similar to 2011 temps, similar year to year
```

## Time Series Decomposition by STL

```{r}
toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf)))

components(toolik_dec) %>% 
  autoplot()

# think about scales of different panels
# original data has scale of 40, trend component range not plotted on same scale range, instead of 5
# but for seasonality, scale is 30, closer to org. data = major contributor
```


